# Statistical Models

**Learning objectives:**

- describe graphs that can help you interpret the results of statistical models, focusing on models that have a single response variable that is either quantitative (a number) or binary (yes/no)

## Correlation plots

- [Correlation coefficients](https://www.statisticshowto.com/probability-and-statistics/correlation-coefficient-formula/) are used to measure how strong a relationship is between two variables. 

- There are several types of correlation coefficient, but the most popular is Pearson’s Product-Moment correlation coefficient, which is commonly used in linear regression. 

- Correlation plots help you to visualize the pairwise relationships between a set of quantitative variables by displaying their correlations using color or shading.

## Saratoga Houses Dataset  

- Let's look at the [Saratoga Houses dataset](https://rkabacoff.github.io/datavis/Data.html#SaratogaHousing), which contains the sale price and characteristics of Saratoga County, NY homes in 2006 in order to explore the relationships among the quantitative variables,

```{r 08-01}
data(SaratogaHouses, package="mosaicData")

# select numeric variables
df <- dplyr::select_if(SaratogaHouses, is.numeric)

# calulate the correlations
r <- cor(df, use="complete.obs")
round(r,2)
```

## [`ggcorrplot`](https://www.rdocumentation.org/packages/ggcorrplot/versions/0.1.1/topics/ggcorrplot) package

- The `ggcorrplot` function in the `ggcorrplot` package can be used to visualize these correlations. 

- By default, it creates a `ggplot2` graph were darker red indicates stronger positive correlations, darker blue indicates stronger negative correlations and white indicates no correlation.

```{r 08-02}
library(ggplot2)
library(ggcorrplot)
ggcorrplot(r)
```

## `ggcorrplot` function

The `ggcorrplot` function has a number of options for customizing the output. For example,

-  `hc.order = TRUE` reorders the variables, placing variables with similar correlation patterns together.

- `type = "lower"` plots the lower portion of the correlation matrix.

- `lab = TRUE` overlays the correlation coefficients (as text) on the plot.

```{r 08-03}
ggcorrplot(r, 
           hc.order = TRUE, 
           type = "lower",
           lab = TRUE)
```


## Linear Regression

- [Linear regression](https://towardsdatascience.com/linear-regression-detailed-view-ea73175f6e86) allows us to explore the relationship between a quantitative response variable and an explanatory variable while other variables are held constant.

- Below is a model to predict home prices, the response variable, by the explanatory variables, lot size ($ft^2$), age (yrs), land value ($\$1000s$), living area ($ft^2$), number of bedrooms and bathrooms and whether the home is on the waterfront or not.

```{r 08-04}
data(SaratogaHouses, package="mosaicData")
houses_lm <- lm(price ~ lotSize + age + landValue +
                  livingArea + bedrooms + bathrooms +
                  waterfront, 
                data = SaratogaHouses)

summary(houses_lm)$coefficients
```

- We estimate that an increase of one square foot of living area is associated with a home price increase of $\$75$, holding the other variables constant. 

- We estimate that a waterfront home costs approximately $\$120,726$ more than non-waterfront home, again controlling for the other variables in the model.

## [`visreg`](http://pbreheny.github.io/visreg/index.html) package 

- The `visreg` package provides tools for visualizing these conditional relationships.

- The `visreg` function takes (1) the model and (2) the variable of interest and plots the conditional relationship, controlling for the other variables. 

- The option `gg = TRUE` is used to produce a `ggplot2` graph.

```{r 08-05}
# conditional plot of price vs. living area
library(ggplot2)
library(visreg)
visreg(houses_lm, "livingArea", gg = TRUE) 
```

- The graph suggests that, holding all else equal, sales price increases with living area in a linear fashion.

## Another linear regression example

- Continuing the example, the price difference between waterfront and non-waterfront homes is plotted, controlling for the other seven variables. 

- Since a `ggplot2` graph is produced, other `ggplot2` functions can be added to customize the graph.

```{r 08-06}
# conditional plot of price vs. waterfront location
visreg(houses_lm, "waterfront", gg = TRUE) +
  scale_y_continuous(label = scales::dollar) +
  labs(title = "Relationship between price and location",
       subtitle = "controlling for lot size, age, land value, bedrooms and bathrooms",
       caption = "source: Saratoga Housing Data (2006)",
       y = "Home Price",
       x = "Waterfront")
```

- We see that there are far fewer homes on the water, and they tend to be more expensive (even controlling for size, age, and land value).

## [Logistic Regression](https://towardsdatascience.com/introduction-to-logistic-regression-66248243c148)

- Logistic regression can be used to explore the relationship between a binary response variable and an explanatory variable while other variables are held constant. 

- Binary response variables have two levels (yes/no, lived/died, pass/fail, malignant/benign). 

- As with linear regression, we can use the `visreg` package to visualize these relationships.

## [CPS85 dataset](https://rkabacoff.github.io/datavis/Data.html#CPS85) and linear regression

- The CPS85 dataset from the `mosaicData` package, contains 1985 data on wages and other characteristics of workers.

- Using the CPS85 data let’s predict the [log-odds](https://towardsdatascience.com/https-towardsdatascience-com-what-and-why-of-log-odds-64ba988bf704) of being married, given one’s sex, age, race and job sector.

```{r 08-07}
# fit logistic model for predicting
# marital status: married/single
data(CPS85, package = "mosaicData")
cps85_glm <- glm(married ~ sex + age + race + sector, 
                 family="binomial", 
                 data=CPS85)
```

## Visualizing logistic regression 

- Using the fitted model, we visualize the relationship between age and the probability of being married, holding the other variables constant. 

- The `scale = "response"` option creates a plot based on a probability (rather than log-odds) scale.

```{r 08-08}
# plot results
library(ggplot2)
library(visreg)
visreg(cps85_glm, "age", 
       gg = TRUE, 
       scale="response") +
  labs(y = "Prob(Married)", 
       x = "Age",
       title = "Relationship of age and marital status",
       subtitle = "controlling for sex, race, and job sector",
       caption = "source: Current Population Survey 1985")
```

- Here we see that the probability of being married is estimated to be roughly 0.5 at age 20 and decreases to 0.1 at age 60, controlling for the other variables.

## Visualizing multiple conditional logistic regression plots

- We can create multiple conditional plots by adding a `by` option. 

- The following code plots the probability of being married given age, seperately for men and women (`by="sex"`), controlling for race and job sector.

```{r 08-09}
# plot results
library(ggplot2)
library(visreg)
visreg(cps85_glm, "age",
       by = "sex",
       gg = TRUE, 
       scale="response") +
  labs(y = "Prob(Married)", 
       x = "Age",
       title = "Relationship of age and marital status",
       subtitle = "controlling for race and job sector",
       caption = "source: Current Population Survey 1985")
```

- In this data, the probability of marriage is very similar for men and women.

## [Survival Analysis](https://towardsdatascience.com/the-complete-introduction-to-survival-analysis-in-python-7523e17737e6)

- In many research settings, especially healthcare research, the response variable is the time to an event i.e. time to recovery, time to death, or time to relapse.

- If the event has not occurred for an observation (either because the study ended or the patient dropped out) the observation is said to be *censored*.

- The [NCCTG Lung Cancer](https://rkabacoff.github.io/datavis/Data.html#Lung) dataset in the `survival` package provides data on the survival times of patients with advanced lung cancer following treatment for up to 34 months. 

- The outcome for each patient is measured by two variables (1) *time*,  survival time in days, (2) *status*, 1 = censored, 2 = dead.

- Thus a patient with time = 305 & status = 2 lived 305 days following treatment. Another patient with time = 400 & status = 1, lived at least 400 days but then dropped out of the study. A patient with time = 1022 & status = 1, survived to the end of the study (34 months).


## Survival Plots using [`ggsurvplot`](https://exts.ggplot2.tidyverse.org/gallery/)

- A survival plot (also called a Kaplan-Meier Curve) can be used to illustrates the probability that an individual survives up to and including time *t*.


```{r 08-10, warning=FALSE, message=FALSE}
# plot survival curve
library(survival)
library(survminer)

data(lung)
sfit <- survfit(Surv(time, status) ~  1, data=lung)
ggsurvplot(sfit,
            title="Kaplan-Meier curve for lung cancer survival") 
```

- Roughly 50% of patients are still alive 300 days post treatment. 

- We can run `summary(sfit)` for more details.


## Comparing survival probabilities

- It is frequently of great interest whether groups of patients have the same survival probabilities. 

- In this graph, the survival curve for men and women are compared.

```{r 08-11}
# plot survival curve for men and women
sfit <- survfit(Surv(time, status) ~  sex, data=lung)
ggsurvplot(sfit, 
           conf.int=TRUE, 
           pval=TRUE,
           legend.labs=c("Male", "Female"), 
           legend.title="Sex",  
           palette=c("cornflowerblue", "indianred3"), 
           title="Kaplan-Meier Curve for lung cancer survival",
           xlab = "Time (days)")
```

- A couple `ggsurvplot` options: (1) `conf.int` provides confidence intervals, (2) `pval` provides a log-rank test comparing the survival curves.

- The p-value (0.0013) provides strong evidence that men and women have different survival probabilities following treatment.

## Mosaic Plots

- Mosaic charts can display the relationship between categorical variables using rectangles whose areas represent the proportion of cases for any given combination of levels.

- He we look at the [Titanic dataset](https://rkabacoff.github.io/datavis/Data.html#Titanic) and visualize the relationship between the three categorical variables in the code below.

```{r 08-12, message=FALSE}
# input data
library(readr)
titanic <- read_csv("data/titanic.csv")

# create a table
tbl <- xtabs(~Survived + Class + Sex, titanic)
ftable(tbl)
```


## Mosaic plots and the `vcd` package

- Although mosaic charts can be created with `ggplot2` using the [`ggmosaic`](https://cran.r-project.org/web/packages/ggmosaic/vignettes/ggmosaic.html) package, the author recommends using the `vcd` package instead, which won’t create `ggplot2` graphs, but provides a more comprehensive approach to visualizing categorical data.

- In a mosaic plot, the size of the tile is proportional to the percentage of cases in that combination of levels. 

- For example, as seen below, more Titanic passengers perished than survived, and those that perished were primarily 3rd class male passengers and male crew (the largest group).

```{r 08-13, message=FALSE}
# create a mosaic plot from the table
library(vcd)
mosaic(tbl, main = "Titanic data")
```


## Adding color to mosaic plot

- In mosaic plots, the color of the tiles can also be used to indicate the degree relationship among the variables.

- If we assume that these three variables are independent, we can examine the residuals (the error between a predicted value and the observed actual value) from the model and shade the tiles to match. 

- In the graph below, dark blue represents more cases than expected given independence, and dark red represents less cases than expected if independence holds.

```{r 08-14}
mosaic(tbl, 
       shade = TRUE,
       legend = TRUE,
       labeling_args = list(set_varnames = c(Sex = "Gender",
                                             Survived = "Survived",
                                             Class = "Passenger Class")),
       set_labels = list(Survived = c("No", "Yes"),
                         Class = c("1st", "2nd", "3rd", "Crew"),
                         Sex = c("F", "M")),
       main = "Titanic data")
```

- We can see that if class, gender, and survival are independent, we are seeing many more male crew perishing, and 1st, 2nd and 3rd class females surviving than would be expected. 

- Conversely, far fewer 1st class passengers (both male and female) died than would be expected by chance, thus the assumption of independence is rejected. 

- NOTE: For complicated tables, labels can easily overlap. See [`labeling_border`](https://www.rdocumentation.org/packages/vcd/versions/1.4-4/topics/labeling_border), for plotting options.

## Resources

- [R Graph Gallery: Correlogram](https://r-graph-gallery.com/correlogram.html)

- [ggplot2 extension gallery](https://exts.ggplot2.tidyverse.org/gallery/)

- The [`lm`](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/lm) function is used to fit linear models. It can be used to carry out regression, single stratum analysis of variance and analysis of covariance (although aov may provide a more convenient interface for these).

- The [`glm`](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/glm) function is used to fit generalized linear models, specified by giving a symbolic description of the linear predictor and a description of the error distribution.


## Meeting Videos {-}

### Cohort 1 {-}

`r knitr::include_url("https://www.youtube.com/embed/AEKNf7p2hYA")`

<details>
<summary> Meeting chat log </summary>

```
00:38:11	Kotomi Oda:	Putting color makes it much easier to read
00:40:24	Tiffany Kollah:	Reacted to "Putting color makes ..." with ❤️
00:41:16	Lydia Gibson:	http://pbreheny.github.io/visreg/index.html
00:42:51	Lydia Gibson:	https://www.rdocumentation.org/packages/glmnet/versions/4.1-7
00:43:25	Kotomi Oda:	yes
```
</details>
