# Other Graphs

**Learning objectives:**

- cover graphs that do not fit neatly elsewhere 

## 3-D Scatterplot

```{r 09-01}
# basic 3-D scatterplot
library(scatterplot3d)
with(mtcars, {
   scatterplot3d(x = disp,
                 y = wt, 
                 z = mpg,
                 main="3-D Scatterplot Example 1")
})
```

## 

```{r 09-02}
library(scatterplot3d)
with(mtcars, {
  scatterplot3d(x = disp, 
                y = wt, 
                z = mpg, 
                # filled blue circles
                color="blue", 
                pch=19, 
                # lines to the horizontal plane
                type = "h",
                main = "3-D Scatterplot Example 2",
                xlab = "Displacement (cu. in.)",
                ylab = "Weight (lb/1000)",
                zlab = "Miles/(US) Gallon")
})
```

## 

```{r 09-03}
library(scatterplot3d)
with(mtcars, {
  s3d <- scatterplot3d(
    x = disp, 
    y = wt, 
    z = mpg,
    color = "blue", 
    pch = 19,      
    type = "h",
    main = "3-D Scatterplot Example 3",
    xlab = "Displacement (cu. in.)",
    ylab = "Weight (lb/1000)",
    zlab = "Miles/(US) Gallon")
  
  # convert 3-D coords to 2D projection
  s3d.coords <- s3d$xyz.convert(disp, wt, mpg) 
  
  # plot text with 50% shrink and place to right of points
  text(s3d.coords$x, 
       s3d.coords$y,   
       labels = row.names(mtcars),  
       cex = .5, 
       pos = 4)
})
```

##

```{r 09-04}
library(scatterplot3d)

# create column indicating point color
mtcars$pcolor[mtcars$cyl == 4] <- "red"
mtcars$pcolor[mtcars$cyl == 6] <- "blue"
mtcars$pcolor[mtcars$cyl == 8] <- "darkgreen"

with(mtcars, {
    s3d <- scatterplot3d(
      x = disp, 
      y = wt, 
      z = mpg,
      color = pcolor, 
      pch = 19, 
      type = "h", 
      lty.hplot = 2, 
      scale.y = .75,
      main = "3-D Scatterplot Example 4",
      xlab = "Displacement (cu. in.)",
      ylab = "Weight (lb/1000)",
      zlab = "Miles/(US) Gallon")
    
     s3d.coords <- s3d$xyz.convert(disp, wt, mpg)
     text(s3d.coords$x, 
          s3d.coords$y, 
          labels = row.names(mtcars), 
          pos = 4, 
          cex = .5)  
     
# add the legend
legend(#location
       "topleft", 
       inset=.05,
       # suppress legend box, shrink text 50%
       bty="n", 
       cex=.5, 
       title="Number of Cylinders",
       c("4", "6", "8"), 
       fill=c("red", "blue", "darkgreen"))
})
```



## Biplots

```{r 09-05}
# create a biplot
# load data
data(mtcars)

# fit a principal components model
fit <- prcomp(x = mtcars, 
              center = TRUE, 
              scale = TRUE)

# plot the results
library(factoextra)
fviz_pca(fit, 
         repel = TRUE, 
         labelsize = 3) + 
  theme_bw() +
  labs(title = "Biplot of mtcars data")
```



## Bubble Charts

```{r 09-06}

# create a bubble plot
data(mtcars)
library(ggplot2)
ggplot(mtcars, 
       aes(x = wt, y = mpg, size = hp)) +
  geom_point()

```

## 

```{r 09-07}
# create a bubble plot with modifications
ggplot(mtcars, 
       aes(x = wt, y = mpg, size = hp)) +
  geom_point(alpha = .5, 
             fill="cornflowerblue", 
             color="black", 
             shape=21) +
  scale_size_continuous(range = c(1, 14)) +
  labs(title = "Auto mileage by weight and horsepower",
       subtitle = "Motor Trend US Magazine (1973-74 models)",
       x = "Weight (1000 lbs)",
       y = "Miles/(US) gallon",
       size = "Gross horsepower") 
```

## Flow Diagrams

- Sankey Diagrams

- Alluvial Diagrams

## Sankey Diagrams

```{r 09-08}

# Load package
library(networkD3)
 
# Load energy projection data
URL <- "https://cdn.rawgit.com/christophergandrud/networkD3/master/JSONdata/energy.json"
Energy <- jsonlite::fromJSON(URL)

# Now we have 2 data frames: a 'links' data frame with 3 columns (from, to, value), and a 'nodes' data frame that gives the name of each node.
head(Energy$links )
```

```{r 09-09}
head(Energy$nodes)
```

```{r 09-10}
# Thus we can plot it
p <- sankeyNetwork(Links = Energy$links, Nodes = Energy$nodes, Source = "source",
              Target = "target", Value = "value", NodeID = "name",
              units = "TWh", fontSize = 12, nodeWidth = 30)
p
```

## Alluvial Diagrams

```{r 09-11, message=FALSE}
# input data
library(readr)
titanic <- read_csv("data/titanic.csv")

# summarize data
library(dplyr)
titanic_table <- titanic %>%
  group_by(Class, Sex, Survived) %>%
  count()

titanic_table$Survived <- factor(titanic_table$Survived, 
                                 levels = c("Yes", "No"))

head(titanic_table)
```


##

```{r 09-12, warning=FALSE}
# create alluvial diagram
library(ggplot2)
library(ggalluvial)

ggplot(titanic_table,
       aes(axis1 = Class,
           axis2 = Survived,
           y = n)) +
  geom_alluvium(aes(fill = Sex)) +
  geom_stratum() +
  geom_text(stat = "stratum", 
            label.strata = TRUE) +
  scale_x_discrete(limits = c("Class", "Survived"),
                   expand = c(.1, .1)) +
  labs(title = "Titanic data",
       subtitle = "stratified by class, sex, and survival",
       y = "Frequency") +
  theme_minimal()
```

##

```{r 09-13, warning=FALSE}
# create alternative alluvial diagram
library(ggplot2)
library(ggalluvial)
ggplot(titanic_table,
       aes(axis1 = Class,
           axis2 = Sex,
           axis3 = Survived,
           y = n)) +
  geom_alluvium(aes(fill = Class)) +
  geom_stratum() +
  geom_text(stat = "stratum", 
            label.strata = TRUE) +
  scale_x_discrete(limits = c("Class", "Sex", "Survived"),
                   expand = c(.1, .1)) +
  scale_fill_viridis_d() +
  labs(title = "Titanic data",
       subtitle = "stratified by class, sex, and survival",
       y = "Frequency") +
  theme_minimal() +
  theme(legend.position = "none") 

```


## Heatmaps

```{r 09-14}
# create a heatmap
data(mtcars)
library(superheat)
superheat(mtcars, scale = TRUE)
```

## Sorted Heatmap

```{r 09-15}
# sorted heat map
superheat(mtcars,
          scale = TRUE,
          left.label.text.size=3,
          bottom.label.text.size=3,
          bottom.label.size = .05,
          row.dendrogram = TRUE )
```

##

```{r 09-16}
# create heatmap for gapminder data (Asia)
library(tidyr)
library(dplyr)

# load data
data(gapminder, package="gapminder")

# subset Asian countries
asia <- gapminder %>%
  filter(continent == "Asia") %>%
  select(year, country, lifeExp)

# convert to long to wide format
plotdata <- spread(asia, year, lifeExp)

# save country as row names
plotdata <- as.data.frame(plotdata)
row.names(plotdata) <- plotdata$country
plotdata$country <- NULL

# row order
sort.order <- order(plotdata$"2007")

# color scheme
library(RColorBrewer)
colors <- rev(brewer.pal(5, "Blues"))


# create the heat map
superheat(plotdata,
          scale = FALSE,
          left.label.text.size=3,
          bottom.label.text.size=3,
          bottom.label.size = .05,
          heat.pal = colors,
          order.rows = sort.order,
          title = "Life Expectancy in Asia")
```

## Radar charts

```{r 09-17, message=FALSE, warning=FALSE}
#devtools::install_github("ricardo-bion/ggradar")

# create a radar chart

# prepare data
data(msleep, package = "ggplot2")
library(ggradar)
library(scales)
library(dplyr)

plotdata <- msleep %>%
  filter(name %in% c("Cow", "Dog", "Pig")) %>%
  select(name, sleep_total, sleep_rem, 
         sleep_cycle, brainwt, bodywt) %>%
  rename(group = name) %>%
  mutate_at(vars(-group),
            funs(rescale))
plotdata

# generate radar chart
ggradar(plotdata, 
        grid.label.size = 4,
        axis.label.size = 4, 
        group.point.size = 5,
        group.line.width = 1.5,
        legend.text.size= 10) +
  labs(title = "Mammals, size, and sleep")
```


## Scaterplot matrix

```{r 09-18, message=FALSE, warning=FALSE}
library(GGally)

# prepare data
data(msleep, package="ggplot2")
library(dplyr)
df <- msleep %>% 
  mutate(log_brainwt = log(brainwt),
         log_bodywt = log(bodywt)) %>%
  select(log_brainwt, log_bodywt, sleep_total, sleep_rem)
 

# create a scatterplot matrix
ggpairs(df)
```

## 

```{r 09-19, warning=FALSE, message=FALSE}

# custom function for density plot
my_density <- function(data, mapping, ...){
  ggplot(data = data, mapping = mapping) + 
    geom_density(alpha = 0.5,
                 fill = "cornflowerblue", ...)
}

# custom function for scatterplot
my_scatter <- function(data, mapping, ...){
  ggplot(data = data, mapping = mapping) + 
    geom_point(alpha = 0.5,
               color = "cornflowerblue") + 
    geom_smooth(method=lm, 
                se=FALSE, ...)
}


# create scatterplot matrix
ggpairs(df, 
        lower=list(continuous = my_scatter), 
        diag = list(continuous = my_density)) +
  labs(title = "Mammal size and sleep characteristics") +
  theme_bw()
```


## Waterfall Charts

```{r 09-20}
# create company income statement
category <- c("Sales", "Services", "Fixed Costs", 
              "Variable Costs", "Taxes")
amount <- c(101000, 52000, -23000, -15000, -10000)
income <- data.frame(category, amount) 

# create waterfall chart
library(ggplot2)
library(waterfalls)
waterfall(income) 
```

##

```{r 09-21}
# create waterfall chart with total column
waterfall(income, 
          calc_total=TRUE, 
          total_axis_text = "Net",
          total_rect_text_color="black",
          total_rect_color="goldenrod1") +
  scale_y_continuous(label=scales::dollar) +
  labs(title = "West Coast Profit and Loss", 
       subtitle = "Year 2017",
       y="", 
       x="") +
  theme_minimal() 
```


## Word clouds


```{r 09-22, warning=FALSE}
#Charge the wordcloud library
library(wordcloud)
 
#Create a list of words (Random words concerning my work)
a <- c("Cereal","WSSMV","SBCMV","Experimentation","Talk","Conference","Writing", 
   "Publication","Analysis","Bioinformatics","Science","Statistics","Data", 
   "Programming","Wheat","Virus","Genotyping","Work","Fun","Surfing","R", "R",
   "Data-Viz","Python","Linux","Programming","Graph Gallery","Biologie", "Resistance",
   "Computing","Data-Science","Reproductible","GitHub","Script")
 
#I give a frequency to each word of this list 
b <- sample(seq(0,1,0.01) , length(a) , replace=TRUE) 
 
#The package will automatically make the wordcloud ! (I add a black background)
par(bg="black") 
wordcloud(a , b , col=terrain.colors(length(a) , alpha=0.9) , rot.per=0.3 )

```

```{r 09-23}
# library
library(wordcloud2) 
 
# have a look to the example dataset
# head(demoFreq)

# Basic plot
wordcloud2(data=demoFreq, size=1.6)
```


## Resources

[Word Clouds:`wordcloud2`](https://cran.r-project.org/web/packages/wordcloud2/vignettes/wordcloud.html)

[r graph gallery: Wordcloud](https://r-graph-gallery.com/wordcloud.html)

[r graph gallery: Sankey Charts](https://r-graph-gallery.com/sankey-diagram.html)

## Meeting Videos {-}

### Cohort 1 {-}

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>
